---
title: "VAR(1) DSEM"
author: "Seungju Kim"
format: 
  html:
    toc: true
    toc-depth: 6
    embed-resources: true
editor_options: 
  chunk_output_type: console
---

This paper demonstrates how to perform a classic discrete-time `Dynamic Structural Equation Models` (DSEMs) using three statistical programs: `Mplus` and `Blimp`. 

# Preparation
## Load Packages
```{r}
library(pacman)
p_load(tidyverse, haven, openesm, rblimp, MplusAutomation, here)
```

## Import Data
```{r}
# org_data <- get_dataset("0037", metadata_version = "1.1.1", version = "1.0.0") 
# saveRDS(org_data, here("data/data.Rds"))
org_data <- readRDS(here("data/data.Rds"))
data <- org_data$data
cite(org_data)
```

:::callout-note
We use data by Fisher and Soster on affect and smoking. This is an open dataset accessible through the R package `openesm`.

Our main focus will be to examine if affect predicts smoking.
:::

## Prep Data
```{r}
data <- data %>%
  group_by(id) %>%
  arrange(id, day, beep) %>%
  mutate(time = row_number()) %>%
  ungroup()

data_dsem <- data %>%
  group_by(id) %>%
  arrange(id, day, beep) %>%
  mutate(time = row_number()) %>%
  ungroup() %>%
  # keep only what we need
  select(id, time, day, beep, craving_smoking, stressed)


data_dsem <- data_dsem |> as.data.frame()

## Recode character id to numeric
data_dsem$id <- data_dsem$id |> as.numeric()
```



# DSEM

## AR(1) 

### Blimp
#### Model Specification
```{r}
#| eval: false
ar_blimp <- rblimp(
    data = data_dsem,
    clusterid = 'id',
    timeid = 'time',
    latent = 'id = stressed_mean craving_mean AR CL',
    model = '
    level2.models:
        1 -> stressed_mean craving_mean AR CL;
    level1.models:
        stressed ~ 1@stressed_mean;
        craving_smoking ~ 
            1@craving_mean
            (craving_smoking.lag - craving_mean)@AR
            (stressed - stressed_mean)@CL;',
    seed = 51823,
    burn = 10000,
    iter = 50000
)

saveRDS(ar_blimp, here("model_objects/ar_blimp.Rds"))
```

#### Results
```{r}
ar_blimp <- readRDS(here("model_objects/ar_blimp.Rds"))
output(ar_blimp)
```




## VAR(1)

### Mplus

#### Model Specification
```{r}
#| eval: false
# # Build and run model
dsem_fit <- mplusModeler(
  mplusObject(
    TITLE = "DSEM - Stressed predicting Craving (random phi and beta);",

    VARIABLE = "
      CLUSTER = id;
      LAGGED = craving_smoking(1);
      TINTERVAL = time(1);",

    ANALYSIS = "
      TYPE = TWOLEVEL RANDOM;
      ESTIMATOR = BAYES;
      BITERATIONS = 20000 (10000);
      CHAINS = 4;
      PROCESSORS = 4;",

    MODEL = "
      %WITHIN%
        phi  | craving_smoking ON craving_smoking&1;
        beta | craving_smoking ON stressed;

      %BETWEEN%
        craving_smoking;
        stressed;
        [phi];
        [beta];
        phi beta craving_smoking stressed WITH
        phi beta craving_smoking stressed;",

    OUTPUT = "TECH1 TECH8 CINTERVAL(HPD);",

    usevariables = c("id", "time", "craving_smoking", "stressed"),
    rdata = data_dsem
  ),
  dataout  = here("model_objects/data_dsem.dat"),
  modelout = here("model_objects/dsem_model.inp"),
  run      = 1L,
  writeData = "ifmissing",
  hashfilename = FALSE
)
```

#### Results
```{r}
# Read results
results <- readModels(here("model_objects/dsem_model.out"), quiet = TRUE)

# Fixed effects: population means of random slopes
results$parameters$unstandardized %>%
  filter(paramHeader == "Means",
         param %in% c("PHI", "BETA")) %>%
  select(param, est, posterior_sd, lower_2.5ci, upper_2.5ci)

# Between-level variances
results$parameters$unstandardized %>%
  filter(paramHeader == "Variances") %>%
  select(param, est, posterior_sd, lower_2.5ci, upper_2.5ci)

# Residual variances (within-level)
results$parameters$unstandardized %>%
  filter(paramHeader == "Residual.Variances") %>%
  select(param, est, posterior_sd, lower_2.5ci, upper_2.5ci)

# Between-level correlations among random effects
results$parameters$unstandardized %>%
  filter(str_detect(paramHeader, "\\.WITH")) %>%
  select(paramHeader, param, est, posterior_sd, lower_2.5ci, upper_2.5ci)

# Convergence: PSR values from TECH8
results$tech8$chain_convergence

# Model summary
results$summaries
```


### Blimp

#### Model Specification
```{r}
#| eval: false
var_blimp <- rblimp(
    data = data_dsem,
    clusterid = 'id',
    timeid = 'time',
    latent = 'id = stressed_mean craving_mean AR_1 CL_1 AR_2 CL_2',
    model = '
    level2.models:
        1 -> stressed_mean craving_mean AR_1 CL_1 AR_2 CL_2;
    level1.models:
        stressed ~ 
            1@stressed_mean
            (stressed.lag - stressed_mean)@AR_1
            (craving_smoking - craving_mean)@CL_1
            ;
        craving_smoking ~ 
            1@craving_mean
            (craving_smoking.lag - craving_mean)@AR_2
            (stressed - stressed_mean)@CL_2;',
    seed = 51823,
    burn = 10000,
    iter = 20000
)

saveRDS(var_blimp, here("model_objects/ar_blimp.Rds"))
```

#### Results
```{r}
var_blimp <- readRDS(here("model_objects/blimp_dsem.Rds"))
output(var_blimp)
```

# RDSEM

## AR(1)

### Blimp
#### Model Specification
```{r}
#| eval: false
ar_1_rdsem_blimp <- rblimp(
    data = data_dsem,
    clusterid = 'id',
    timeid = 'time',
    latent = 'id = stressed_mean craving_mean craving_trend AR CL',
    model = '
    ## SPECIFY VARIABLES ##
    # here we create a lag variable + time trend
    lag_hat = craving_mean + (time - 1)*craving_trend;
    
    # here we create the lag_res which calculates change from the 
    # prior time point after takign into account time trend
    lag_res = ifelse(time <= 1, 0, craving_smoking.lag - lag_hat);
    
    level2.models:
        1 -> stressed_mean craving_mean AR CL craving_trend;
    
    level1.models:
        stressed ~ 1@stressed_mean;
        craving_smoking ~ 
            1@craving_mean
            lag_res@AR
            (stressed - stressed_mean)@CL
            time@craving_trend
    ',
    seed = 51823,
    burn = 10000,
    iter = 50000
)

saveRDS(ar_rdsem_blimp, here("model_objects/ar_rdsem_blimp.Rds"))
```

#### Results
```{r}
ar_rdsem_blimp <- readRDS(here("model_objects/ar_rdsem_blimp.Rds"))
output(ar_rdsem_blimp)
```

## VAR(1)

### Blimp
#### Model Specification
```{r}
#| eval: false
var_rdsem_blimp <- rblimp(
    data = data_dsem,
    clusterid = 'id',
    timeid = 'time',
    latent = 'id = craving_mean stressed_mean 
              CL_1 AR_1 CL_2 AR_2
              craving_trend stressed_trend',
    model = '
    ## DETRENDED LAGS FOR CRAVING ##
    craving_lag_hat = craving_mean + (time - 1)*craving_trend;
    craving_lag_res = ifelse(time <= 1, 0, 
                        craving_smoking.lag - craving_lag_hat);
    
    ## DETRENDED LAGS FOR STRESSED ##
    stressed_lag_hat = stressed_mean + (time - 1)*stressed_trend;
    stressed_lag_res = ifelse(time <= 1, 0, 
                        stressed.lag - stressed_lag_hat);
    
    level2.models:
        1 -> craving_mean stressed_mean 
             CL_1 AR_1 CL_2 AR_2
             craving_trend stressed_trend;
    
    level1.models:
        stressed ~ 
            1@stressed_mean
            craving_lag_res@CL_1
            stressed_lag_res@AR_1
            time@stressed_trend;
        
        craving_smoking ~ 
            1@craving_mean
            craving_lag_res@AR_2
            stressed_lag_res@CL_2
            time@craving_trend;',
    seed = 51823,
    burn = 10000,
    iter = 50000
)

saveRDS(var_rdsem_blimp, here("model_objects/var_rdsem_blimp.Rds"))
```

#### Results
```{r}
var_rdsem_blimp <- readRDS(here("model_objects/var_rdsem_blimp.Rds"))
output(var_rdsem_blimp)
```
