---
title: "Comparing Dynamic Structural Equation Modeling with Mplus and Blimp: 
author: 
  - "Seungju Kim"
  - "Peter J. Jankowski"
format: yeah
  html:
    toc: true
    toc-depth: 6
    embed-resources: true
editor_options: 
  chunk_output_type: console
---

This paper demonstrates how to perform a classic discrete-time `Dynamic Structural Equation Models` (DSEMs) using three statistical programs: `Mplus` and `Blimp`. 

# Preparation
## Load Packages
```{r}
library(pacman)
p_load(tidyverse, haven, openesm, rblimp, MplusAutomation, here)
```

## Import Data
```{r}
# org_data <- get_dataset("0037", metadata_version = "1.1.1", version = "1.0.0") 
# saveRDS(org_data, here("data/data.Rds"))
org_data <- readRDS(here("data/data.Rds"))
data <- org_data$data
cite(org_data)
```

:::callout-note
We use data by Fisher and Soster on affect and smoking. This is an open dataset accessible through the R package `openesm`.

Our main focus will be to examine if affect predicts smoking.
:::

## Prep Data
```{r}
data <- data %>%
  group_by(id) %>%
  arrange(id, day, beep) %>%
  mutate(time = row_number()) %>%
  ungroup()

data_dsem <- data %>%
  group_by(id) %>%
  arrange(id, day, beep) %>%
  mutate(time = row_number()) %>%
  ungroup() %>%
  # keep only what we need
  select(id, time, day, beep, craving_smoking, stressed)


data_dsem <- data_dsem |> as.data.frame()

## Recode character id to numeric
data_dsem$id <- data_dsem$id |> as.numeric()
```


# Empirical Example

## DSEM AR(1) 
### Blimp
#### Model Specification
```{r}
#| eval: false
ar_blimp <- rblimp(
    data = data_dsem,
    clusterid = 'id',
    timeid = 'time',
    latent = 'id = stressed_mean craving_mean AR CL',
    model = '
    level2.models:
        1 -> stressed_mean craving_mean AR CL;
    level1.models:
        stressed ~ 1@stressed_mean;
        craving_smoking ~ 
            1@craving_mean
            (craving_smoking.lag - craving_mean)@AR
            (stressed - stressed_mean)@CL;',
    seed = 51823,
    burn = 10000,
    iter = 50000
)

saveRDS(ar_blimp, here("model_objects/ar_blimp.Rds"))
```

#### Results
```{r}
ar_blimp <- readRDS(here("model_objects/ar_blimp.Rds"))
output(ar_blimp)
```




## DSEM VAR(1)

### Mplus

#### Model Specification
```{r}
#| eval: false
# # Build and run model
dsem_fit <- mplusModeler(
  mplusObject(
    TITLE = "DSEM - Stressed predicting Craving (random phi and beta);",

    VARIABLE = "
      CLUSTER = id;
      LAGGED = craving_smoking(1);
      TINTERVAL = time(1);",

    ANALYSIS = "
      TYPE = TWOLEVEL RANDOM;
      ESTIMATOR = BAYES;
      BITERATIONS = 20000 (10000);
      CHAINS = 4;
      PROCESSORS = 4;",

    MODEL = "
      %WITHIN%
        phi  | craving_smoking ON craving_smoking&1;
        beta | craving_smoking ON stressed;

      %BETWEEN%
        craving_smoking;
        stressed;
        [phi];
        [beta];
        phi beta craving_smoking stressed WITH
        phi beta craving_smoking stressed;",

    OUTPUT = "TECH1 TECH8 CINTERVAL(HPD);",

    usevariables = c("id", "time", "craving_smoking", "stressed"),
    rdata = data_dsem
  ),
  dataout  = here("model_objects/data_dsem.dat"),
  modelout = here("model_objects/dsem_model.inp"),
  run      = 1L,
  writeData = "ifmissing",
  hashfilename = FALSE
)
```

#### Results
```{r}
# Read results
results <- readModels(here("model_objects/dsem_model.out"), quiet = TRUE)

# Fixed effects: population means of random slopes
results$parameters$unstandardized %>%
  filter(paramHeader == "Means",
         param %in% c("PHI", "BETA")) %>%
  select(param, est, posterior_sd, lower_2.5ci, upper_2.5ci)

# Between-level variances
results$parameters$unstandardized %>%
  filter(paramHeader == "Variances") %>%
  select(param, est, posterior_sd, lower_2.5ci, upper_2.5ci)

# Residual variances (within-level)
results$parameters$unstandardized %>%
  filter(paramHeader == "Residual.Variances") %>%
  select(param, est, posterior_sd, lower_2.5ci, upper_2.5ci)

# Between-level correlations among random effects
results$parameters$unstandardized %>%
  filter(str_detect(paramHeader, "\\.WITH")) %>%
  select(paramHeader, param, est, posterior_sd, lower_2.5ci, upper_2.5ci)

# Convergence: PSR values from TECH8
results$tech8$chain_convergence

# Model summary
results$summaries
```


### Blimp

#### Model Specification
```{r}
#| eval: false
var_blimp <- rblimp(
    data = data_dsem,
    clusterid = 'id',
    timeid = 'time',
    latent = 'id = stressed_mean craving_mean AR_1 CL_1 AR_2 CL_2',
    model = '
    level2.models:
        1 -> stressed_mean craving_mean AR_1 CL_1 AR_2 CL_2;
    level1.models:
        stressed ~ 
            1@stressed_mean
            (stressed.lag - stressed_mean)@AR_1
            (craving_smoking - craving_mean)@CL_1
            ;
        craving_smoking ~ 
            1@craving_mean
            (craving_smoking.lag - craving_mean)@AR_2
            (stressed - stressed_mean)@CL_2;',
    seed = 51823,
    burn = 10000,
    iter = 20000
)

saveRDS(var_blimp, here("model_objects/ar_blimp.Rds"))
```

#### Results
```{r}
var_blimp <- readRDS(here("model_objects/blimp_dsem.Rds"))
output(var_blimp)
```

## RDSEM AR(1)

### Blimp
#### Model Specification
```{r}
#| eval: false
ar_1_rdsem_blimp <- rblimp(
    data = data_dsem,
    clusterid = 'id',
    timeid = 'time',
    latent = 'id = stressed_mean craving_mean craving_trend AR CL',
    model = '
    ## SPECIFY VARIABLES ##
    # here we create a lag variable + time trend
    lag_hat = craving_mean + (time - 1)*craving_trend;
    
    # here we create the lag_res which calculates change from the 
    # prior time point after takign into account time trend
    lag_res = ifelse(time <= 1, 0, craving_smoking.lag - lag_hat);
    
    level2.models:
        1 -> stressed_mean craving_mean AR CL craving_trend;
    
    level1.models:
        stressed ~ 1@stressed_mean;
        craving_smoking ~ 
            1@craving_mean
            lag_res@AR
            (stressed - stressed_mean)@CL
            time@craving_trend
    ',
    seed = 51823,
    burn = 10000,
    iter = 50000
)

saveRDS(ar_rdsem_blimp, here("model_objects/ar_rdsem_blimp.Rds"))
```

#### Results
```{r}
ar_rdsem_blimp <- readRDS(here("model_objects/ar_rdsem_blimp.Rds"))
output(ar_rdsem_blimp)
```

## RDSEM VAR(1)

### Blimp
#### Model Specification
```{r}
#| eval: false
var_rdsem_blimp <- rblimp(
    data = data_dsem,
    clusterid = 'id',
    timeid = 'time',
    latent = 'id = craving_mean stressed_mean 
              CL_1 AR_1 CL_2 AR_2
              craving_trend stressed_trend',
    model = '
    ## DETRENDED LAGS FOR CRAVING ##
    craving_lag_hat = craving_mean + (time - 1)*craving_trend;
    craving_lag_res = ifelse(time <= 1, 0, 
                        craving_smoking.lag - craving_lag_hat);
    
    ## DETRENDED LAGS FOR STRESSED ##
    stressed_lag_hat = stressed_mean + (time - 1)*stressed_trend;
    stressed_lag_res = ifelse(time <= 1, 0, 
                        stressed.lag - stressed_lag_hat);
    
    level2.models:
        1 -> craving_mean stressed_mean 
             CL_1 AR_1 CL_2 AR_2
             craving_trend stressed_trend;
    
    level1.models:
        stressed ~ 
            1@stressed_mean
            craving_lag_res@CL_1
            stressed_lag_res@AR_1
            time@stressed_trend;
        
        craving_smoking ~ 
            1@craving_mean
            craving_lag_res@AR_2
            stressed_lag_res@CL_2
            time@craving_trend;',
    seed = 51823,
    burn = 10000,
    iter = 50000
)

saveRDS(var_rdsem_blimp, here("model_objects/var_rdsem_blimp.Rds"))
```

#### Results
```{r}
var_rdsem_blimp <- readRDS(here("model_objects/var_rdsem_blimp.Rds"))
output(var_rdsem_blimp)
```

# Simulation Study

y_it = α_i + φ_i*(y_i,t-1 - α_i) + β_i*(x_it - μx_i) + u_it
x_it = μx_i + v_it

## Create Dataset DSEM AR(1)

### Population Specifications
```{r}
# ============================================
# SHARED PARAMETERS (all four simulations)
# ============================================
craving_mu       <- 45    # craving person mean
stressed_mu      <- 50    # stressed person mean
craving_between  <- 230   # between-person var in craving mean
stressed_between <- 225   # between-person var in stressed mean
craving_resid    <- 635   # within-person residual var (craving)
stressed_resid   <- 545   # within-person residual var (stressed)

# ============================================
# SIMULATION 1: AR(1) DSEM
# ============================================
AR_mean  <- 0.30   # rounded up from 0.086 for moderate effect
CL_mean  <- 0.10   # rounded from 0.108
AR_var   <- 0.01
CL_var   <- 0.02

# ============================================
# SIMULATION 2: VAR(1) DSEM
# Empirical: AR_1=0.262, AR_2=0.069, CL_1=0.021, CL_2=0.074
# ============================================
AR_1_mean  <- 0.30   # stressed AR (rounded up for moderate)
AR_2_mean  <- 0.10   # craving AR
CL_1_mean  <- 0.05   # craving -> stressed (rounded from 0.021)
CL_2_mean  <- 0.10   # stressed -> craving (rounded from 0.074)
AR_1_var   <- 0.02
AR_2_var   <- 0.01
CL_1_var   <- 0.002
CL_2_var   <- 0.001
stressed_resid_var2 <- 485  # from VAR(1) RDSEM output

# ============================================
# SIMULATION 3: AR(1) RDSEM
# Adds linear trend to craving
# Empirical trend: 0.017, rounded to moderate
# ============================================
AR_mean_r    <- 0.30
CL_mean_r    <- 0.10
AR_var_r     <- 0.01
CL_var_r     <- 0.02
trend_mean   <- 0.10   # clear moderate trend per occasion
trend_var    <- 0.01

# ============================================
# SIMULATION 4: VAR(1) RDSEM
# Adds linear trend to both craving and stressed
# ============================================
AR_1_mean_r       <- 0.30
AR_2_mean_r       <- 0.10
CL_1_mean_r       <- 0.05
CL_2_mean_r       <- 0.10
craving_trend_mean  <- 0.10
stressed_trend_mean <- 0.10
craving_trend_var   <- 0.01
stressed_trend_var  <- 0.01
```


### Generate Data

#### AR DSEM
```{r}
# ============================================
# FUNCTION 1: AR(1) DSEM
# craving ~ lag(craving) + stressed
# stressed is purely exogenous (no dynamics)
# ============================================

generate_ar1_dsem <- function(N, T, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  
  # Population parameters
  AR_mean  <- 0.30
  CL_mean  <- 0.10
  AR_var   <- 0.01
  CL_var   <- 0.02
  
  # Draw person-specific parameters
  AR_i            <- rnorm(N, AR_mean, sqrt(AR_var))
  CL_i            <- rnorm(N, CL_mean, sqrt(CL_var))
  craving_mean_i  <- rnorm(N, craving_mu,  sqrt(craving_between))
  stressed_mean_i <- rnorm(N, stressed_mu, sqrt(stressed_between))
  
  data_list <- vector("list", N)
  
  for (i in 1:N) {
    craving  <- numeric(T)
    stressed <- numeric(T)
    
    # Initialize
    craving[1]  <- craving_mean_i[i] + 
                   rnorm(1, 0, sqrt(craving_resid))
    stressed[1] <- stressed_mean_i[i] + 
                   rnorm(1, 0, sqrt(stressed_resid))
    
    for (t in 2:T) {
      # stressed is purely exogenous — no AR structure
      stressed[t] <- stressed_mean_i[i] + 
                     rnorm(1, 0, sqrt(stressed_resid))
      
      # craving has AR and cross-lagged from stressed
      craving[t]  <- craving_mean_i[i] +
                     AR_i[i] * (craving[t-1]  - craving_mean_i[i]) +
                     CL_i[i] * (stressed[t]   - stressed_mean_i[i]) +
                     rnorm(1, 0, sqrt(craving_resid))
    }
    
    data_list[[i]] <- data.frame(
      id              = i,
      time            = 1:T,
      craving_smoking = craving,
      stressed        = stressed
    )
  }
  
  do.call(rbind, data_list)
}
```

#### VAR DSEM
```{r}
# ============================================
# FUNCTION 2: VAR(1) DSEM
# craving ~ lag(craving) + lag(stressed)
# stressed ~ lag(stressed) + lag(craving)
# both outcomes have dynamics
# ============================================

generate_var1_dsem <- function(N, T, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  
  # Population parameters
  AR_1_mean <- 0.30   # stressed AR
  AR_2_mean <- 0.10   # craving AR
  CL_1_mean <- 0.05   # craving -> stressed
  CL_2_mean <- 0.10   # stressed -> craving
  AR_1_var  <- 0.02
  AR_2_var  <- 0.01
  CL_1_var  <- 0.002
  CL_2_var  <- 0.001
  stressed_resid_var <- 485
  
  # Draw person-specific parameters
  AR_1_i          <- rnorm(N, AR_1_mean, sqrt(AR_1_var))
  AR_2_i          <- rnorm(N, AR_2_mean, sqrt(AR_2_var))
  CL_1_i          <- rnorm(N, CL_1_mean, sqrt(CL_1_var))
  CL_2_i          <- rnorm(N, CL_2_mean, sqrt(CL_2_var))
  craving_mean_i  <- rnorm(N, craving_mu,  sqrt(craving_between))
  stressed_mean_i <- rnorm(N, stressed_mu, sqrt(stressed_between))
  
  data_list <- vector("list", N)
  
  for (i in 1:N) {
    craving  <- numeric(T)
    stressed <- numeric(T)
    
    # Initialize at person means
    craving[1]  <- craving_mean_i[i] + 
                   rnorm(1, 0, sqrt(craving_resid))
    stressed[1] <- stressed_mean_i[i] + 
                   rnorm(1, 0, sqrt(stressed_resid_var))
    
    for (t in 2:T) {
      # Both variables now have AR and cross-lagged structure
      # Using lagged values for both (true VAR structure)
      craving[t]  <- craving_mean_i[i] +
                     AR_2_i[i] * (craving[t-1]  - craving_mean_i[i]) +
                     CL_2_i[i] * (stressed[t-1] - stressed_mean_i[i]) +
                     rnorm(1, 0, sqrt(craving_resid))
      
      stressed[t] <- stressed_mean_i[i] +
                     AR_1_i[i] * (stressed[t-1] - stressed_mean_i[i]) +
                     CL_1_i[i] * (craving[t-1]  - craving_mean_i[i]) +
                     rnorm(1, 0, sqrt(stressed_resid_var))
    }
    
    data_list[[i]] <- data.frame(
      id              = i,
      time            = 1:T,
      craving_smoking = craving,
      stressed        = stressed
    )
  }
  
  do.call(rbind, data_list)
}
```

#### AR RSEM
```{r}
# ============================================
# FUNCTION 3: AR(1) RDSEM
# Same as AR(1) DSEM but craving has a 
# linear time trend
# ============================================

generate_ar1_rdsem <- function(N, T, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  
  # Population parameters
  AR_mean      <- 0.30
  CL_mean      <- 0.10
  AR_var       <- 0.01
  CL_var       <- 0.02
  trend_mean   <- 0.10   # clear moderate trend per occasion
  trend_var    <- 0.01
  
  # Draw person-specific parameters
  AR_i            <- rnorm(N, AR_mean,    sqrt(AR_var))
  CL_i            <- rnorm(N, CL_mean,    sqrt(CL_var))
  trend_i         <- rnorm(N, trend_mean, sqrt(trend_var))
  craving_mean_i  <- rnorm(N, craving_mu,  sqrt(craving_between))
  stressed_mean_i <- rnorm(N, stressed_mu, sqrt(stressed_between))
  
  data_list <- vector("list", N)
  
  for (i in 1:N) {
    craving  <- numeric(T)
    stressed <- numeric(T)
    
    # Initialize
    craving[1]  <- craving_mean_i[i] + 
                   trend_i[i] * (1 - 1) +   # trend starts at 0 at t=1
                   rnorm(1, 0, sqrt(craving_resid))
    stressed[1] <- stressed_mean_i[i] + 
                   rnorm(1, 0, sqrt(stressed_resid))
    
    for (t in 2:T) {
      # stressed remains purely exogenous
      stressed[t] <- stressed_mean_i[i] + 
                     rnorm(1, 0, sqrt(stressed_resid))
      
      # predicted value of craving at t-1 (for detrending the lag)
      lag_hat <- craving_mean_i[i] + trend_i[i] * (t - 2)
      lag_res <- craving[t-1] - lag_hat
      
      # craving has AR (detrended), cross-lagged, and trend
      craving[t]  <- craving_mean_i[i] +
                     trend_i[i] * (t - 1) +
                     AR_i[i] * lag_res +
                     CL_i[i] * (stressed[t] - stressed_mean_i[i]) +
                     rnorm(1, 0, sqrt(craving_resid))
    }
    
    data_list[[i]] <- data.frame(
      id              = i,
      time            = 1:T,
      craving_smoking = craving,
      stressed        = stressed
    )
  }
  
  do.call(rbind, data_list)
}
```

#### VAR RDSEM
```{r}
# ============================================
# FUNCTION 4: VAR(1) RDSEM
# Both craving and stressed have AR, 
# cross-lagged effects, AND linear time trends
# ============================================

generate_var1_rdsem <- function(N, T, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  
  # Population parameters
  AR_1_mean           <- 0.30
  AR_2_mean           <- 0.10
  CL_1_mean           <- 0.05
  CL_2_mean           <- 0.10
  AR_1_var            <- 0.02
  AR_2_var            <- 0.01
  CL_1_var            <- 0.002
  CL_2_var            <- 0.001
  craving_trend_mean  <- 0.10
  stressed_trend_mean <- 0.10
  craving_trend_var   <- 0.01
  stressed_trend_var  <- 0.01
  stressed_resid_var  <- 485
  
  # Draw person-specific parameters
  AR_1_i               <- rnorm(N, AR_1_mean, sqrt(AR_1_var))
  AR_2_i               <- rnorm(N, AR_2_mean, sqrt(AR_2_var))
  CL_1_i               <- rnorm(N, CL_1_mean, sqrt(CL_1_var))
  CL_2_i               <- rnorm(N, CL_2_mean, sqrt(CL_2_var))
  craving_trend_i      <- rnorm(N, craving_trend_mean,  sqrt(craving_trend_var))
  stressed_trend_i     <- rnorm(N, stressed_trend_mean, sqrt(stressed_trend_var))
  craving_mean_i       <- rnorm(N, craving_mu,  sqrt(craving_between))
  stressed_mean_i      <- rnorm(N, stressed_mu, sqrt(stressed_between))
  
  data_list <- vector("list", N)
  
  for (i in 1:N) {
    craving  <- numeric(T)
    stressed <- numeric(T)
    
    # Initialize at t=1 (trend term is 0 at t=1)
    craving[1]  <- craving_mean_i[i] + 
                   rnorm(1, 0, sqrt(craving_resid))
    stressed[1] <- stressed_mean_i[i] + 
                   rnorm(1, 0, sqrt(stressed_resid_var))
    
    for (t in 2:T) {
      # Detrended lagged residuals for both variables
      craving_lag_hat  <- craving_mean_i[i]  + craving_trend_i[i]  * (t - 2)
      stressed_lag_hat <- stressed_mean_i[i] + stressed_trend_i[i] * (t - 2)
      
      craving_lag_res  <- craving[t-1]  - craving_lag_hat
      stressed_lag_res <- stressed[t-1] - stressed_lag_hat
      
      # craving: AR (detrended) + CL from stressed (detrended) + trend
      craving[t]  <- craving_mean_i[i] +
                     craving_trend_i[i]  * (t - 1) +
                     AR_2_i[i] * craving_lag_res +
                     CL_2_i[i] * stressed_lag_res +
                     rnorm(1, 0, sqrt(craving_resid))
      
      # stressed: AR (detrended) + CL from craving (detrended) + trend
      stressed[t] <- stressed_mean_i[i] +
                     stressed_trend_i[i] * (t - 1) +
                     AR_1_i[i] * stressed_lag_res +
                     CL_1_i[i] * craving_lag_res +
                     rnorm(1, 0, sqrt(stressed_resid_var))
    }
    
    data_list[[i]] <- data.frame(
      id              = i,
      time            = 1:T,
      craving_smoking = craving,
      stressed        = stressed
    )
  }
  
  do.call(rbind, data_list)
}
```

### Test
```{r}
# ============================================
# SANITY CHECKS
# ============================================

set.seed(123)
test1 <- generate_ar1_dsem(N = 100, T = 50)
test2 <- generate_var1_dsem(N = 100, T = 50)
test3 <- generate_ar1_rdsem(N = 100, T = 50)
test4 <- generate_var1_rdsem(N = 100, T = 50)

# ---- CHECK 1: Person-level means ----
# Should be centered around craving=45, stressed=50

check_means <- function(dat, label) {
  dat %>%
    group_by(id) %>%
    summarise(
      m_craving  = mean(craving_smoking),
      m_stressed = mean(stressed)
    ) %>%
    summarise(
      across(everything(), mean)
    ) %>%
    mutate(model = label)
}

bind_rows(
  check_means(test1, "AR1 DSEM"),
  check_means(test2, "VAR1 DSEM"),
  check_means(test3, "AR1 RDSEM"),
  check_means(test4, "VAR1 RDSEM")
)

# ---- CHECK 2: Within-person variances ----
# craving residual should be ~635, stressed ~545

check_vars <- function(dat, label) {
  dat %>%
    group_by(id) %>%
    summarise(
      var_craving  = var(craving_smoking),
      var_stressed = var(stressed)
    ) %>%
    summarise(
      across(everything(), mean)
    ) %>%
    mutate(model = label)
}

bind_rows(
  check_vars(test1, "AR1 DSEM"),
  check_vars(test2, "VAR1 DSEM"),
  check_vars(test3, "AR1 RDSEM"),
  check_vars(test4, "VAR1 RDSEM")
)

# ---- CHECK 3: Trend is visible in RDSEM data ----
# Should show clear upward slope in craving over time

p1 <- test3 %>%
  group_by(time) %>%
  summarise(m_craving = mean(craving_smoking)) %>%
  ggplot(aes(time, m_craving)) +
  geom_line() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "AR(1) RDSEM: Mean craving over time",
       subtitle = "Should show upward linear trend (slope ~0.10)") +
  theme_minimal()

p2 <- test4 %>%
  group_by(time) %>%
  summarise(
    m_craving  = mean(craving_smoking),
    m_stressed = mean(stressed)
  ) %>%
  pivot_longer(-time) %>%
  ggplot(aes(time, value, color = name)) +
  geom_line() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "VAR(1) RDSEM: Mean trajectories over time",
       subtitle = "Both should show upward trend (slope ~0.10)") +
  theme_minimal()

p1
p2

# ---- CHECK 4: No trend in DSEM data ----
# Craving should be flat over time

test1 %>%
  group_by(time) %>%
  summarise(m_craving = mean(craving_smoking)) %>%
  ggplot(aes(time, m_craving)) +
  geom_line() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "AR(1) DSEM: Mean craving over time",
       subtitle = "Should be flat — no trend") +
  theme_minimal()


# Check 4b: VAR(1) DSEM should also be flat
test2 %>%
  group_by(time) %>%
  summarise(
    m_craving  = mean(craving_smoking),
    m_stressed = mean(stressed)
  ) %>%
  pivot_longer(-time) %>%
  ggplot(aes(time, value, color = name)) +
  geom_line() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "VAR(1) DSEM: Mean trajectories over time",
       subtitle = "Both should be flat — no trend") +
  theme_minimal()

# ---- CHECK 5: Verify AR structure ----
# Correlation between craving[t] and craving[t-1] should be ~0.30

check_ar <- function(dat, label) {
  dat %>%
    group_by(id) %>%
    arrange(time) %>%
    summarise(
      ar_cor = cor(
        craving_smoking[-n()],           # craving t
        craving_smoking[-1],             # craving t+1
        use = "complete.obs"
      )
    ) %>%
    summarise(mean_ar = mean(ar_cor)) %>%
    mutate(model = label)
}

bind_rows(
  check_ar(test1, "AR1 DSEM"),
  check_ar(test2, "VAR1 DSEM"),
  check_ar(test3, "AR1 RDSEM"),
  check_ar(test4, "VAR1 RDSEM")
)
```

## Blimp
### True Values
```{r}
# ============================================
# TRUTH VALUES FOR EACH SIMULATION
# ============================================

truth_ar1_dsem <- list(
  AR_mean  = 0.30,
  CL_mean  = 0.10,
  AR_var   = 0.01,
  CL_var   = 0.02
)

truth_var1_dsem <- list(
  AR_1_mean = 0.30,
  AR_2_mean = 0.10,
  CL_1_mean = 0.05,
  CL_2_mean = 0.10,
  AR_1_var  = 0.02,
  AR_2_var  = 0.01,
  CL_1_var  = 0.002,
  CL_2_var  = 0.001
)

truth_ar1_rdsem <- list(
  AR_mean    = 0.30,
  CL_mean    = 0.10,
  trend_mean = 0.10,
  AR_var     = 0.01,
  CL_var     = 0.02,
  trend_var  = 0.01
)

truth_var1_rdsem <- list(
  AR_1_mean          = 0.30,
  AR_2_mean          = 0.10,
  CL_1_mean          = 0.05,
  CL_2_mean          = 0.10,
  craving_trend_mean = 0.10,
  stressed_trend_mean = 0.10,
  AR_1_var           = 0.02,
  AR_2_var           = 0.01,
  CL_1_var           = 0.002,
  CL_2_var           = 0.001,
  craving_trend_var  = 0.01,
  stressed_trend_var = 0.01
)
```

### Extraction
```{r}
# ============================================
# BLIMP EXTRACTION HELPERS
# ============================================
# These pull the posterior median and 95% CI
# from the rblimp estimates table

extract_blimp_ar1_dsem <- function(fit) {
  est <- fit@estimates
  
  # Helper to grab median and CI for a parameter by name
  get_param <- function(pattern) {
    row <- est[grep(pattern, est$Parameter, ignore.case = TRUE), ]
    if (nrow(row) == 0) return(c(est = NA, lo = NA, hi = NA))
    c(est = row$Estimate[1], lo = row$`2.5%`[1], hi = row$`97.5%`[1])
  }
  
  tibble(
    AR_mean  = get_param("AR.*Intercept|Intercept.*AR")[["est"]],
    AR_mean_lo = get_param("AR.*Intercept|Intercept.*AR")[["lo"]],
    AR_mean_hi = get_param("AR.*Intercept|Intercept.*AR")[["hi"]],
    
    CL_mean  = get_param("CL.*Intercept|Intercept.*CL")[["est"]],
    CL_mean_lo = get_param("CL.*Intercept|Intercept.*CL")[["lo"]],
    CL_mean_hi = get_param("CL.*Intercept|Intercept.*CL")[["hi"]],
    
    AR_var   = get_param("AR.*Residual Var|Residual Var.*AR")[["est"]],
    CL_var   = get_param("CL.*Residual Var|Residual Var.*CL")[["est"]]
  )
}

extract_blimp_var1_dsem <- function(fit) {
  est <- fit@estimates
  
  get_param <- function(pattern) {
    row <- est[grep(pattern, est$Parameter, ignore.case = TRUE), ]
    if (nrow(row) == 0) return(c(est = NA, lo = NA, hi = NA))
    c(est = row$Estimate[1], lo = row$`2.5%`[1], hi = row$`97.5%`[1])
  }
  
  tibble(
    AR_1_mean    = get_param("AR_1.*Intercept")[["est"]],
    AR_1_mean_lo = get_param("AR_1.*Intercept")[["lo"]],
    AR_1_mean_hi = get_param("AR_1.*Intercept")[["hi"]],
    
    AR_2_mean    = get_param("AR_2.*Intercept")[["est"]],
    AR_2_mean_lo = get_param("AR_2.*Intercept")[["lo"]],
    AR_2_mean_hi = get_param("AR_2.*Intercept")[["hi"]],
    
    CL_1_mean    = get_param("CL_1.*Intercept")[["est"]],
    CL_1_mean_lo = get_param("CL_1.*Intercept")[["lo"]],
    CL_1_mean_hi = get_param("CL_1.*Intercept")[["hi"]],
    
    CL_2_mean    = get_param("CL_2.*Intercept")[["est"]],
    CL_2_mean_lo = get_param("CL_2.*Intercept")[["lo"]],
    CL_2_mean_hi = get_param("CL_2.*Intercept")[["hi"]],
    
    AR_1_var  = get_param("AR_1.*Residual Var")[["est"]],
    AR_2_var  = get_param("AR_2.*Residual Var")[["est"]],
    CL_1_var  = get_param("CL_1.*Residual Var")[["est"]],
    CL_2_var  = get_param("CL_2.*Residual Var")[["est"]]
  )
}

extract_blimp_ar1_rdsem <- function(fit) {
  est <- fit@estimates
  
  get_param <- function(pattern) {
    row <- est[grep(pattern, est$Parameter, ignore.case = TRUE), ]
    if (nrow(row) == 0) return(c(est = NA, lo = NA, hi = NA))
    c(est = row$Estimate[1], lo = row$`2.5%`[1], hi = row$`97.5%`[1])
  }
  
  tibble(
    AR_mean      = get_param("AR.*Intercept")[["est"]],
    AR_mean_lo   = get_param("AR.*Intercept")[["lo"]],
    AR_mean_hi   = get_param("AR.*Intercept")[["hi"]],
    
    CL_mean      = get_param("CL.*Intercept")[["est"]],
    CL_mean_lo   = get_param("CL.*Intercept")[["lo"]],
    CL_mean_hi   = get_param("CL.*Intercept")[["hi"]],
    
    trend_mean   = get_param("craving_trend.*Intercept")[["est"]],
    trend_mean_lo = get_param("craving_trend.*Intercept")[["lo"]],
    trend_mean_hi = get_param("craving_trend.*Intercept")[["hi"]],
    
    AR_var    = get_param("AR.*Residual Var")[["est"]],
    CL_var    = get_param("CL.*Residual Var")[["est"]],
    trend_var = get_param("craving_trend.*Residual Var")[["est"]]
  )
}

extract_blimp_var1_rdsem <- function(fit) {
  est <- fit@estimates
  
  get_param <- function(pattern) {
    row <- est[grep(pattern, est$Parameter, ignore.case = TRUE), ]
    if (nrow(row) == 0) return(c(est = NA, lo = NA, hi = NA))
    c(est = row$Estimate[1], lo = row$`2.5%`[1], hi = row$`97.5%`[1])
  }
  
  tibble(
    AR_1_mean           = get_param("AR_1.*Intercept")[["est"]],
    AR_1_mean_lo        = get_param("AR_1.*Intercept")[["lo"]],
    AR_1_mean_hi        = get_param("AR_1.*Intercept")[["hi"]],
    
    AR_2_mean           = get_param("AR_2.*Intercept")[["est"]],
    AR_2_mean_lo        = get_param("AR_2.*Intercept")[["lo"]],
    AR_2_mean_hi        = get_param("AR_2.*Intercept")[["hi"]],
    
    CL_1_mean           = get_param("CL_1.*Intercept")[["est"]],
    CL_1_mean_lo        = get_param("CL_1.*Intercept")[["lo"]],
    CL_1_mean_hi        = get_param("CL_1.*Intercept")[["hi"]],
    
    CL_2_mean           = get_param("CL_2.*Intercept")[["est"]],
    CL_2_mean_lo        = get_param("CL_2.*Intercept")[["lo"]],
    CL_2_mean_hi        = get_param("CL_2.*Intercept")[["hi"]],
    
    craving_trend_mean    = get_param("craving_trend.*Intercept")[["est"]],
    craving_trend_mean_lo = get_param("craving_trend.*Intercept")[["lo"]],
    craving_trend_mean_hi = get_param("craving_trend.*Intercept")[["hi"]],
    
    stressed_trend_mean    = get_param("stressed_trend.*Intercept")[["est"]],
    stressed_trend_mean_lo = get_param("stressed_trend.*Intercept")[["lo"]],
    stressed_trend_mean_hi = get_param("stressed_trend.*Intercept")[["hi"]],
    
    AR_1_var           = get_param("AR_1.*Residual Var")[["est"]],
    AR_2_var           = get_param("AR_2.*Residual Var")[["est"]],
    CL_1_var           = get_param("CL_1.*Residual Var")[["est"]],
    CL_2_var           = get_param("CL_2.*Residual Var")[["est"]],
    craving_trend_var  = get_param("craving_trend.*Residual Var")[["est"]],
    stressed_trend_var = get_param("stressed_trend.*Residual Var")[["est"]]
  )
}
```


### Results
```{r}
source(here::here("simulation_run.R"))  # load fresh code
plan(multisession, workers = 8)     # start new workers with updated code
# plan(sequential)

results_sim1 <- run_simulation_blimp_parallel(sim_num = 1, 
                                              N_vec = c(50, 100, 200), 
                                              T_vec = c(30, 50), 
                                              n_reps = 1000, 
                                              burn = 2000, 
                                              iter = 5000)

results_sim2 <- run_simulation_blimp_parallel(sim_num = 2, 
                                              N_vec = c(50, 100, 200), 
                                              T_vec = c(30, 50), 
                                              n_reps = 1000, 
                                              burn = 2000, 
                                              iter = 5000)
results_sim3 <- run_simulation_blimp_parallel(sim_num = 3, 
                                              N_vec = c(50, 100, 200), 
                                              T_vec = c(30, 50), 
                                              n_reps = 1000, 
                                              burn = 2000, 
                                              iter = 5000)
results_sim4 <- run_simulation_blimp_parallel(sim_num = 4, 
                                              N_vec = c(50, 100, 200), 
                                              T_vec = c(30, 50), 
                                              n_reps = 1000, 
                                              burn = 2000, 
                                              iter = 5000)
```




