---
title: "VAR(1) DSEM"
author: "Seungju Kim"
format: 
  html:
    toc: true
    toc-depth: 6
    embed-resources: true
editor_options: 
  chunk_output_type: console
---

This paper demonstrates how to perform a classic discrete-time `Dynamic Structural Equation Models` (DSEMs) using three statistical programs: `Mplus`, `Blimp`, and `mlts in R`. 

# Preparation
## Load Packages
```{r}
library(pacman)
p_load(tidyverse, ctsem, haven, openesm, rblimp, MplusAutomation, here, mlts)
```

## Import Data
```{r}
org_data <- get_dataset("0037", metadata_version = "1.1.1", version = "1.0.0") 
data <- org_data$data
cite(org_data)
```

:::callout-note
We use data by Fisher and Soster on affect and smoking. This is an open dataset accessible through the R package `openesm`.

Our main focus will be to examine if affect predicts smoking.
:::

## Prep Data
```{r}
data <- data %>%
  group_by(id) %>%
  arrange(id, day, beep) %>%
  mutate(time = row_number()) %>%
  ungroup()

data_dsem <- data %>%
  group_by(id) %>%
  arrange(id, day, beep) %>%
  mutate(time = row_number()) %>%
  ungroup() %>%
  # keep only what we need
  select(id, time, day, beep, craving_smoking, stressed)


data_dsem <- data_dsem |> as.data.frame()

## Recode character id to numeric
data_dsem$id <- data_dsem$id |> as.numeric()
```

# AR(1) DSEM

# VAR(1) DSEM

## Mplus

### Model Specification
```{r}
#| eval: false
# # Build and run model
dsem_fit <- mplusModeler(
  mplusObject(
    TITLE = "DSEM - Stressed predicting Craving (random phi and beta);",

    VARIABLE = "
      CLUSTER = id;
      LAGGED = craving_smoking(1);
      TINTERVAL = time(1);",

    ANALYSIS = "
      TYPE = TWOLEVEL RANDOM;
      ESTIMATOR = BAYES;
      BITERATIONS = 20000 (10000);
      CHAINS = 4;
      PROCESSORS = 4;",

    MODEL = "
      %WITHIN%
        phi  | craving_smoking ON craving_smoking&1;
        beta | craving_smoking ON stressed;

      %BETWEEN%
        craving_smoking;
        stressed;
        [phi];
        [beta];
        phi beta craving_smoking stressed WITH
        phi beta craving_smoking stressed;",

    OUTPUT = "TECH1 TECH8 CINTERVAL(HPD);",

    usevariables = c("id", "time", "craving_smoking", "stressed"),
    rdata = data_dsem
  ),
  dataout  = here("model_objects/data_dsem.dat"),
  modelout = here("model_objects/dsem_model.inp"),
  run      = 1L,
  writeData = "ifmissing",
  hashfilename = FALSE
)
```

### Results
```{r}
# Read results
results <- readModels(here("model_objects/dsem_model.out"), quiet = TRUE)

# Fixed effects: population means of random slopes
results$parameters$unstandardized %>%
  filter(paramHeader == "Means",
         param %in% c("PHI", "BETA")) %>%
  select(param, est, posterior_sd, lower_2.5ci, upper_2.5ci)

# Between-level variances
results$parameters$unstandardized %>%
  filter(paramHeader == "Variances") %>%
  select(param, est, posterior_sd, lower_2.5ci, upper_2.5ci)

# Residual variances (within-level)
results$parameters$unstandardized %>%
  filter(paramHeader == "Residual.Variances") %>%
  select(param, est, posterior_sd, lower_2.5ci, upper_2.5ci)

# Between-level correlations among random effects
results$parameters$unstandardized %>%
  filter(str_detect(paramHeader, "\\.WITH")) %>%
  select(paramHeader, param, est, posterior_sd, lower_2.5ci, upper_2.5ci)

# Convergence: PSR values from TECH8
results$tech8$chain_convergence

# Model summary
results$summaries
```


## rBlimp

### Model Specification
```{r}
#| eval: false
mymodel <- rblimp(
    data = data_dsem,
    clusterid = 'id',
    timeid = 'time',
    latent = 'id = stressed_mean_j craving_mean_j phi_j beta_j',
    model = '
    level2.models:
        # by default blimp specifies an intercept
        #   when you use the @ with a latent variable
        # Correlate the random effects
        craving_mean_j phi_j beta_j ~~ craving_mean_j phi_j beta_j;
    level1.models:
        stressed ~ 1@stressed_mean_j;
        craving_smoking ~ 1@craving_mean_j
            (craving_smoking.lag - craving_mean_j)@phi_j
            (stressed - stressed_mean_j)@beta_j;',
    seed = 51823,
    burn = 10000,
    iter = 20000
)

saveRDS(mymodel, here("model_objects/blimp_dsem.Rds"))
```

### Results
```{r}
mymodel <- readRDS(here("model_objects/blimp_dsem.Rds"))
output(mymodel)
```

## mlts

### Model Specification (1/2)
```{r}
VAR1mod <- mlts_model(
  q = 2,
  max_lag = 1,
  inno_covs_zero = TRUE,
  is_exogenous = 2 
)
```

#### Examime Formula
```{r}
#| eval: false
mlts_model_formula(VAR1mod, file = here("visual/var1_mlts.pdf"))
```

:::callout-note
See the PDF file generated in the `visual` folder.
:::

#### Visualize Model
```{r}
mlts_paths(
  VAR1mod,
  width = 14,
  height = 10,
  fig_margins.x = c(0, 10),
  fig_margins.y = c(1, 10)
)
```

:::callout-tip
A very nice feature in `mlts` is the ability to visualize your model, which can aid interpretation and specification for applied researchers.
:::

#### Rescale Data

```{r}
# Rescale to -3 to 3
# rescale_to_range <- function(x, new_min = -3, new_max = 3) {
#   obs_min <- min(x, na.rm = TRUE)
#   obs_max <- max(x, na.rm = TRUE)
#   if (obs_min == obs_max) return(rep(0, length(x)))  # handle zero-variance case
#   (x - obs_min) / (obs_max - obs_min) * (new_max - new_min) + new_min
# }
# 
# data_dsem <- data_dsem %>%
#   mutate(
#     craving_scaled = rescale_to_range(craving_smoking),
#     stressed_scaled = rescale_to_range(stressed)
#   )
```

### Results
```{r}
#| eval: false
fit_mlts <- mlts_fit(
  model   = VAR1mod,
  data    = data_dsem_clean,
  id      = "id",
  ts      = c("craving_smoking", "stressed"),
  time    = "time",
  tinterval = 1,
  iter    = 10000,
  chains  = 2,
  cores   = 6
)

saveRDS(fit_mlts, here("model_objects/mlts_fit.Rds"))
```

```{r}
fit_mlts <- readRDS(here("model_objects/mlts_fit.Rds"))
summary(fit_mlts)
```

# RDSEM
